---
title: "Mapping with an 800 Pound Gorilla"
author: "Jason Preszler"
date: 2019-03-18
output: html_document
tags: ["Python", "matplotlib", "map"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I've been focusing on `python` recently to become a bi-lingual data scientist. Probably my least favorite thing about `python` is its plotting libraries - there are too many options built on top of `matplotlib` which pre-dates `pandas` dataframes. This makes for some clunky code and blurry boundaries (both "is that a `seaborn`, `pandas`, or `matplotlib` function?" and situations with 3 equally messy solutions but in very different ways). In my opinion, `ggplot2`'s deep interplay with dataframes makes a lot more sense and `ggplot`'s layers make it easy to change plot type (just switch the `geom_`), add facets, and tweak aesthetics. I think this [post](https://dsaber.com/2016/10/02/a-dramatic-tour-through-pythons-data-visualization-landscape-including-ggplot-and-altair/) does a great job comparing the main `python` plotting libraries and also illustrates why `matplotlib` turns me off a bit - too many loops!

Despite `matplotlib`'s issues, it can make some really nice plots! Today I'll be doing a little mapping so show off it's prettier side.

## Data
The data I'll be using is of the places I've lived - GPS coordinates of 9 cities in 7 states (for security reasons I'm omitting where I was born). I'm going to use `R`'s `maps` package to get most of the data (because it's soooo easy).

```{r, eval=FALSE}
library(dplyr)
library(ggplot2)
library(maps)

cities <- filter(maps::us.cities, name %in% c("Reno NV", "Sacramento CA", "Salt Lake City UT", "Tacoma WA", "Boise ID", "Tulare CA", "Harrisonburg VA", "Charlotte NC", "Seattle WA"))
```

This dataframe was saved to a `.csv` for later use. The rest will be in `python` to draw a pretty map, overlay some points and connect the points with great circles.

## Basemap
There are two main ways I know of to draw maps in `python`: `matplotlib`'s `basemap` toolkit and `geopandas`. As you might expect, `geopandas` is generally a lot nicer to use, but can be limiting. I'm going to use `basemap` since most of the data I need to plot is just to get the background map drawn and `basemap` has that built in (no shapefiles or extra datasets). 

First, we'll load our `python` libraries and set-up the underlying map.
```{python}
import pandas as pd
import matplotlib.pyplot as plt
import matplotlib as mpl
from mpl_toolkits.basemap import Basemap

m = Basemap(llcrnrlon=-130.,llcrnrlat=20.,urcrnrlon=-60. , urcrnrlat=50., rsphere=(6378137.00,6356752.3142),
            resolution='l',projection='merc',
            lat_0=35.,lon_0=-95.)
m.drawcoastlines()
m.drawstates()
m.drawcountries()
m.fillcontinents()
plt.title("The US")
plt.show()
```

The `Basemap` command has many options, here I've set the lower-left corner and upper-right corner, the Mercator projection, low resolution, and a center of $35^{\circ}N$, $95^{\circ}W$. This will include the first 48 states of the US roughly centered. Then we add coastlines, states, countries, and fill the land. These last commands are what makes `Basemap` nice - I don't need shapefiles or other data to get all this detail onto the map.

## Adding Points
Now I can add points for the cities. First, I'll load the dataframe of the gps data and build lists of the order in which I moved.
```{python}
city = pd.read_csv('../../static/files/city-gps2.csv')
print(city.head())

#lists of longitude and latitude in order.
clon = [city.long[3], city.long[8], city.long[4], 
        city.long[7], city.long[5], city.long[1], 
        city.long[6], city.long[7], city.long[2], 
        city.long[0]]
clat = [city.lat[3], city.lat[8], city.lat[4], 
        city.lat[7], city.lat[5], city.lat[1], 
        city.lat[6], city.lat[7], city.lat[2], 
        city.lat[0]]
```

Next, we add these points (`(clon,clat)` pairs) to the earlier basemap.
```{python}
x,y = m(clon, clat)
m.scatter(x,y,marker='*', color='r', zorder=5) #the zorder is important
plt.show()
```

## Connecting the Points
Now we can connect the points. If I were really bored, I could trace out the actual paths driven to get from one place to another. I'm not, so I'm just going to use `Basemap`'s `drawgreatcircle()` method to connect points with a great circle. This is why having the lists of latitude and longitude was really done. 

```{python}
for i in range(0,len(clon)-1):
    m.drawgreatcircle(clon[i],clat[i],clon[i+1],clat[i+1],linewidth=4,color='b')
plt.title("Places Lived with Great Circle Moves")
plt.show()
```

Not bad, but adding an `alpha` setting to change transparency based on the order of moves helps. The earliest move will be at $1/3$ (light, but visible) while the most recent will be $1$, with a linear increment. I'm also going to close the old plot and put all the mapping code together here.
```{python}
plt.close() 


m = Basemap(llcrnrlon=-130.,llcrnrlat=20.,urcrnrlon=-60. , urcrnrlat=50., rsphere=(6378137.00,6356752.3142),
            resolution='l',projection='merc',
            lat_0=35.,lon_0=-95.)
m.drawcoastlines()
m.drawstates()
m.drawcountries()
m.shadedrelief()
x,y = m(clon, clat)
m.scatter(x,y,marker='*', color='r', zorder=5)
for i in range(0,len(clon)-1):
    m.drawgreatcircle(clon[i],clat[i],clon[i+1],clat[i+1],linewidth=4,color='b', alpha=((1/3)+i*(2/27)))
plt.title("Places Lived with Great Circle Moves")
plt.show()
```

You may have noticed that I used `.shadedrelief()` instead of `.fillcontinents()`. `Basemap` has several similar options to provide some real quality to maps. Here are the `.etopo()` and `.bluemarble()` options as well:

```{python}
plt.close()


plt.subplot(1,2,1)
m2 = Basemap(llcrnrlon=-130.,llcrnrlat=20.,urcrnrlon=-60. , urcrnrlat=50., rsphere=(6378137.00,6356752.3142),
            resolution='l',projection='merc',
            lat_0=35.,lon_0=-95.)
m2.drawcoastlines()
m2.drawstates()
m2.drawcountries()
m2.bluemarble()
x,y = m2(clon, clat)
m2.scatter(x,y,marker='*', color='r', zorder=5)
for i in range(0,len(clon)-1):
    m2.drawgreatcircle(clon[i],clat[i],clon[i+1],clat[i+1],linewidth=4,color='b', alpha=(.3+i*(2/27)))

plt.subplot(1,2,2)
m3 = Basemap(llcrnrlon=-130.,llcrnrlat=20.,urcrnrlon=-60. , urcrnrlat=50., rsphere=(6378137.00,6356752.3142),
            resolution='l',projection='merc',
            lat_0=35.,lon_0=-95.)
m3.drawcoastlines()
m3.drawstates()
m3.drawcountries()
m3.etopo()
x,y = m3(clon, clat)
m3.scatter(x,y,marker='*', color='r', zorder=5)
for i in range(0,len(clon)-1):
    m3.drawgreatcircle(clon[i],clat[i],clon[i+1],clat[i+1],linewidth=4,color='b', alpha=(.3+i*(2/27)))

plt.show()
```

It's no `ggplot2`, but very nice for an 800 pound gorilla[*](https://dsaber.com/2016/10/02/a-dramatic-tour-through-pythons-data-visualization-landscape-including-ggplot-and-altair/).