---
title: "Tidy Clouds"
author: "Jason Preszler"
date: 2019-01-11
output: html_document
tags: ["R","wordcloud","gutenbergr", "shiny"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

In my data visualization class I had the students get a book from [Project Gutenberg](https://www.gutenberg.org) using the `gutenbergr` package and build a word cloud using `tidytext` and `wordcloud`. It's much easier that the "old" corpus/text mapping approach, and when the students were sharing their clouds they started showing the cloud and having students try to guess the book. This made me think of using a Shiny runtime to make a little word cloud guessing game.

# Building Clouds
First, here's the code to grab a book and build a word cloud with line-by-line comments but the code is pretty straight forward (especially if you've build a word cloud before).

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

library(gutenbergr)
library(tidytext)
library(wordcloud)
library(RColorBrewer)

cloud_from_book <- function(gutenberg_id){
  #download book using gutenbergr
  book <- gutenberg_download(gutenberg_id)
  #turn line per row into word per row
  Words <- unnest_tokens(book, word, text)
  #remove stop_words using the standard english stop_words
  WordsReduced <- anti_join(Words, stop_words)
  #turn word list into frequency table
  WRCount <- WordsReduced %>% count(word) %>% ungroup()
  #build wordcloud
  wordcloud(WRCount$word, WRCount$n, random.order = FALSE, max.words = 75, colors = brewer.pal(8, "Dark2"))
}
```
Let's test it out on a whale of a tale:
```{r}
cloud_from_book(2489)
```

# Shiny Clouds
To make this more interesting, we'll turn this into an embedded shiny app. It will have a pretty simple design, two buttons: "New Book" and "Show Book Info". Selecting "New Book" will update the wordcloud from a random gutenberg_id, while "Show Book Info" will reveal title and other book information.

Here's the `ui` function which is pretty basic:
```{r ui, eval=FALSE}
ui <- fluidPage(
   titlePanel("Random Book Word Cloud"),
   sidebarLayout(
      sidebarPanel(
        actionButton("new", "New Book"),
        br(),
        actionButton("reveal", "Show Book Info")
      ),
      
      mainPanel(
         plotOutput("distPlot"),
         tableOutput("BookInfo"))
   )
)
```

And now the server function (assuming our `cloud_from_book` function is already defined). Tis is a little more complicated because of the reactive values to control the table of book info:
```{r server, eval=FALSE}
server <- function(input, output) {
  
   v<- reactiveValues(random_id=NULL, metaTab=NULL)
   
   observeEvent(input$new,{
     v$random_id<-sample(gutenberg_metadata$gutenberg_id, 1)
     v$metaTab <- NULL
   })
   observeEvent(input$reveal, {
     v$metaTab<-filter(gutenberg_metadata, gutenberg_id==v$random_id)[,2:3]
   })
   
   output$distPlot <- renderPlot({
     if(!is.null(v$random_id)){
       cloud_from_book(v$random_id)
     }
   })

   output$BookInfo <- renderTable({
     if(!is.null(v$metaTab)){v$metaTab}
   })
}
```

The `reactiveValues` function produces of list-like structure (the documentation words, not mine...) for reactive variables. Both the `random_id` and `metaTab` are `NULL`. When the `New Book` button is pressed, `random_id` is filled and `metaTab` is set to `NULL` (this resets the table of book info if it's been shown). The `Show Book Info` button (aka `input$reveal`) doesn't do anything to the `random_id` but fills the `metaTab` table. The `if` statements in the `render*` functions then control what is rendered and avoids error messages being printed.

# Embedded [App](https://jpreszler.shinyapps.io/Gutenberg-Clouds/)

```{r app}
knitr::include_app("https://jpreszler.shinyapps.io/Gutenberg-Clouds/", height = "800px")